---
title: 'Project 1:  Reproducible Research'
author: "Austin L. Bistline"
date: "September 10, 2017"
output: 
      html_document:
      keep_md: true
---

## Loading and preprocessing the da
1. Load the data

```{r}
steps.df <- read.csv("activity.csv", header = TRUE, stringsAsFactors = FALSE)
head(steps.df)
```

2. Process/transform the data into a useable format - removing rows containing NA's

```{r}
steps.NA.rm <- na.omit(steps.df)
head(steps.NA.rm)
```

## What is the mean TOTAL number of steps taken per day?

1. Calculate the number of steps taken per day

```{r}
total.steps.per.day <- with(steps.NA.rm, tapply(steps, date, sum))
total.steps.per.day
```

2. Make a histogram of the total number of steps taken each day

```{r}
hist(total.steps.per.day, breaks = length(total.steps.per.day),
     main = "Histogram of Total Steps Each Day",
     xlab = "Steps per Day", ylab = "Number of Days", col = "orange")
```

3. Calculate and report the mean and median of the total number of steps taken each day

```{r}
mean.steps <- mean(total.steps.per.day)
paste("The mean number of steps is ", mean.steps)
```
```{r}
median.steps <- median(total.steps.per.day)
paste("The median number of steps is ", median.steps)
```


## What is the Average Daily Activity Pattern?

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
mean.steps.per.interval <- with(steps.NA.rm, tapply(steps, interval, mean))
plot(unique(steps.NA.rm[, 3]), mean.steps.per.interval, type = "l",
     main = "Steps taken during each 5-minute interval",
     xlab = "Time (24 Hour Clock)", ylab = "Steps")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
time.of.max <- steps.NA.rm[which.max(mean.steps.per.interval), 3]
print(paste("On average, the greatest number of steps occurs at ", time.of.max, "(24-hour time-format)"))
```

The maximum steps occur around 8:35 AM interpreted from [1] above. 

## Inputing missing values

Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}
paste("There are a total of", nrow(steps.df) - nrow(steps.NA.rm), "missing values out of", nrow(steps.df), "data points recorded")
paste("This is ", round((nrow(steps.df)-nrow(steps.NA.rm))/nrow(steps.df)*100, 2), "% of total values.")

```


2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

Strategy:
i. Modify above code to store average values for each 5-minute interval in a new data frame called "ls.of.averages".  
ii. Detect rows where NA's occur in column 1 of the original data frame, "steps.df", and replace them with the average value from "ls.of.averages" corresponding to that particular 5-minute interval from column 3 of the same row in "steps.df". 

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
ls.of.averages <- data.frame(interval = unique(steps.NA.rm[, 3]), 
                             steps = mean.steps.per.interval)
head(ls.of.averages)
```

```{r}
# Initialize a copy dataframe of steps.df
complete.steps.df <- steps.df

# Replace each NA value in column 1 with the rounded integer average of the respective 5-minute interval
for (i in 1:nrow(steps.df)){
      if (is.na(steps.df[i, 1]))
          complete.steps.df[i, 1] <- 
                  as.integer(round(ls.of.averages[match(steps.df[i, 3], ls.of.averages[, 1]), 2]),0)
}
# Demonstrate that there are no longer NA's present
summary(complete.steps.df)
```

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of inputing missing data on the estimates of the total daily number of steps?

Calculating total steps per day from the complete steps dataframe.
```{r}
complete.total.steps.per.day <- with(complete.steps.df, tapply(steps, date, sum))
complete.total.steps.per.day
```

Creating the new histogram.

```{r}
hist(complete.total.steps.per.day, breaks = length(complete.total.steps.per.day),
     main = "Histogram of Complete Total Steps Each Day",
     xlab = "Steps per Day", ylab = "Number of Days", col = "orange")
```

Calculating the mean of the complete steps dataframe.

```{r}
mean.complete.steps <- mean(complete.total.steps.per.day)
paste("The mean number of steps of the complete steps dataframe is ", round(mean.complete.steps, 2))
paste("This is a change of ", abs(mean.complete.steps-mean.steps), "from before NA's were replaced.")
```

Calculating the median of the complete steps dataframe.

```{r}
median.complete.steps <- median(complete.total.steps.per.day)
paste("The median number of steps of the complete steps datafrme is ", median.complete.steps)
paste("This is a change of ", abs(median.complete.steps-median.steps))
```

The impact of replacing missing values in the steps dataframe is very minimal with regards to the mean and median values, however, it had a significant impact on the number of days with the median number of steps, increasing from 7 to 11 days, observable in the histogram.  

# Are there differences in activity patterns between weekdays and weekends?

### For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.

1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}
library(dplyr)
day = weekdays(as.Date(steps.df[, 2]))
day.factor = rep(0, length(day))
for (i in 1:length(day)){
      if (day[i] == "Saturday" || day[i] == "Sunday") {
            day.factor[i] = "weekend"
      } else {
            day.factor[i] = "weekday"
      }
}

complete.steps.df <- mutate(complete.steps.df, day = as.factor(day.factor))
```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.

```{r}
library(lattice)
weekday.steps <- complete.steps.df[complete.steps.df$day == "weekday", ]
weekend.steps <- complete.steps.df[complete.steps.df$day == "weekend", ]
weekday.average <- round(with(weekday.steps, tapply(steps, interval, mean)))
weekend.average <- round(with(weekend.steps, tapply(steps, interval, mean)))
average.steps = c(weekday.average, weekend.average)
interval = as.numeric(rep(unique(steps.df[, 3]), 2))
days = c(rep("weekday", 288), rep("weekend", 288))
average.df = as.data.frame(cbind(average.steps, interval, as.factor(days)))


xyplot(average.steps ~ interval|days, data = average.df, main="Average Steps per Interval (Weekend vs. Weekday)", xlab="Time (24-hour format)", ylab = "Average Number of Steps", layout=c(1,2), type = "l")
```




